Global patterns of forest autotrophic carbon fluxes

### 2020-05-03 manuscript/manuscript draft.Rmd
worked on writing, references
I'm no longer able to knit the document.

```{r setup, include = FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/becky/Dropbox (Smithsonian)/GitHub/Global_Productivity/manuscript/')
# knitr::opts_knit$set(root.dir = '/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/Global_Productivity/manuscript/')
```

@@

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity, while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients (**REFS**) [@piao_forest_2010]. Critically, FACF also define forest interactions with Earth's changing climate. Over 69 Gt of CO~2~ cycle through Earth's forests each year [@badgley_terrestrial_2019], and in recent decades their net C sequestration (~2.4 Gt C yr^-1^) offset roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate [@cavaleri_urgent_2015; @rogelj_mitigation_2018]. However, our ability to draw general macroscopic conclusions regarding global variation in FACF with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables; and typically consider only one or a few variables at a time.

@@

The recent development of a global forest carbon database synthesizing multiple variables and including records of stand history (ForC; [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018]) opens up the possibility for a standardized analysis of global scale variation in multiple FACF and the principle climatic drivers of these patterns. In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1). ** see [issue #48 ](https://github.com/forc-db/Global_Productivity/issues/48)** First, we ask how forest autotrophic carbon fluxes (FACF) vary with latitude. We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables. Finally, we consider the relationship between FACF and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on FACF standardized by growing season length. We use a comprehensive global database of forest carbon fluxes to address the above questions for nine carbon fluxes, allowing for an in-depth exploration of the effect of climate on FACF globally.

@@

Variation in allocation to component carbon fluxes along latitudinal gradients was explored for the following pairings: $GPP:NPP$, $ANPP:BNPP$, and $ANPP_{foliage}:ANPP_{woody.stem}$; $ANPP_{foliage}:NPP$, $ANPP_{woody.stem}:NPP$, $ANPP:NPP$, and $BNPP:NPP$. For each set of paired fluxes, measurements taken at the same site and plot, and in the same year, were paired together, and the ratio of each pair of measurements calculated. The ratios were regressed against latitude and climate variables, using the linear model specified above. Cook's distance analyses were carried out for each of the models, and indicated that data from a few high-elevation sites were having a disproportionate influence on the regressions. To account for this, models were re-run using only data from sites $\le$ 1000m.

@@

In total, we analyzed 1228 records from nine FACF variables taken from forests that had experienced no major anthropogenic disturbances within the past 100 years. These records represented a total of 154 distinct geographic areas (Fig. 1, Table 2), across all forested biogeographic and climate zones.

@@

All major carbon fluxes decreased linearly with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes--particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 254, p<0.0001), 50% in NPP (n = 114, p<0.0001) and 45% in ANPP (n = 259, p<0.0001). The FACF that were most poorly predicted by latitude were $BNPP_{root-fine}$ ($R^2$=0.17) and $ANPP_{woody.stem}$=0.18). For all FACF, the relationship with latitude was best predicted by the linear model).

@@

All FACF increased linearly with MAT, and we found no support for a saturation point of FACF with MAT (all p<0.05; Figs. 3-4, S4-S5, Table S2). As with latitude, MAT tended to explain more variation in the larger FACF ($GPP$, $NPP$, $ANPP$, $R_{auto}$) and $ANPP_foliage$ (all $R^2$> 0.4) than in subsidiary and belowground fluxes ($ANPP_{woody-stem}$, $R_{root}$, $BNPP_{root-fine}$; all $R^2$< 0.25).

@@

All fluxes, with the exception of $R_{root}$, showed a positive linear relationship with solar radiation (Figs. S4-S5, Table S2). Solar radiation explained a low proportion of variability in all FACF, explaining less than 20% of the variation in each flux, with the exception of$R_{auto}$ ($R^2$ = 0.26, p<0.05).

@@

Temperature seasonality was a significant predictor of FACF. $GPP$, $NPP$, $ANPP$, and $R_{root}$ exhibited a polynomial relationship with seasonality (all p<0.05; Figs. 4e, S6-7; Table S2). $ANPP_{foliage}$, $ANPP_{woody-stem}$ and $R_{auto}$ decreased linearly with temperature seasonality (all p<0.05; Figs. 4e, S6-S7; Table S2). Temperature seasonality was strongly correlated with annual temperature range, which was likewise a similarly strong predictor of FACF (Table S2).FACF were highest where temperature seasonality = 0, and at an annual temperature range of 15$^\circ$C or lower.

@@

When FACF were standardized by growing season length, correlations with growing season climate--including temperature, precipitation, solar radiation, and PET--were generally weak (Figs. S8-S9). Speficifally, $ANPP$ increased with growing season temperature ($R^2$ = 0.10, p<0.001) and precipitation ($R^2$ = 0.04, p<0.05). Similaryly, $ANPP_{foliage}$ increased slightly with growing season temperature ($R^2$ = 0.16, p<0.01) and precipitation ($R^2$ = 0.09, p<0.05). Growing season solar radiation had a positive influence on $GPP$ ($R^2$ = 0.21, p<0.001), $NPP$ ($R^2$ = 0.21, p<0.001), $BNPP$ ($R^2$ = 0.16, p<0.001) and $BNPP_{fine.root}$ ($R^2$ = 0.12, p<0.01). Growing season PET had a positive influence on $GPP$ ($R^2$ = 0.15, p<0.01), $NPP$ ($R^2$ = 0.18, p<0.01), $BNPP$ ($R^2$ = 0.23, p<0.0001), $BNPP_{fine.root}$ ($R^2$ = 0.11, p<0.05), and $ANPP_{woody-stem}$ ($R^2$ = 0.06, p<0.05). {**Becky, please verify/ edit the following:** There were no other signifcant correlations between growing season length-standardized FACF (9 variables in Table 2) and growing season climate (**which variables?**)}.

@@

Our analysis of a large global database (ForC) reveals how autotrophic carbon fluxes in mature forests vary with latitude and climate on a global scale. We show that, across all nine FACF analyzed, C cycling decreases continually with latitude (*H1.1*; Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals that of tropical forests [@huston_global_2009]. The FACF generally increase in proportion to one another (*H1.2*), with few differences in allocation detectable at this global scale (Fig. S2) and with component fluxes summing appropriately to larger fluxes (Fig. 2), indicating no major, systematic omissions or overestimations of flux components. However, climate explained lower proportions of variability among subsidiary C fluxes (*e.g.*, $ANPP_{woody}$, $BNPP_{fine.root}$, $R_{auto-root}$; Fig. 2; Table S2). Latitudianal variation in FACF is primarily attributable to temperature-related variables (*H3, H4*), particularly MAT (Figs. 3-4). Water availability is also influential, but generally of secondary importance across the range represented in our database (Figs. 3-4). Temperature seasonality and growing season length are closely correlated with MAT and are strong predictors of FACF (*H4*; Figs. 4e-f, S2, S6-S7), though growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of FACF (*H5*; Fig. S9; Table S4). These findings clarify the big picture of how FACF vary with latitude and climate on a global scale.

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*; **REFS**)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands [@potapov_mapping_2008; @poulter_global_2018], so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because C allocation varies with stand age [@delucia_forest_2007] (**See Nobby's comment in manuscript-draft_NK.pdf**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration [@kunert_understanding_2019].
[*purpose for respiration/ other compenents* (**See Nobby's comment in manuscript-draft_NK.pdf**)].

**This paragraph needs work--including re-assessment of results, interpretation, writing**
We show that FACF are broadly consistent in their responses to climate drivers on the global scale (with the exception of some differences in MAT-MAP interactions; Fig. 3), with no pronounced trends in C allocation among the variable pairs tested (Figs. 2, S3). Although variation in allocation has been observed along gradients of elevation [@moser_elevation_2011] and water availability [@newman_above-_2006]--along with non-climatic axes of stand age [@litton_carbon_2007], nutrient availability [@litton_carbon_2007; @gill_belowground_2016], and forest structure [@taylor_greater_2019]--little variation in relation to climate is apparent at the global scale within ForC, which contains the bulk of relevant data. Our conclusion, then, is that hypothesized gradients in allocation along global climate gradients cannot currently be supported for mature forests, although data quantity and standardization is currently insufficent to rule out the possibility that such trends exist. 
**bbl: remove following sentence?** Of particular interest and significance are the relationships amongst $GPP$, net primary productivity ($NPP$ and its components, particularly $ANPP_{woody-stem}$), and respiration ($R_{auto}$ and components). There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age [@delucia_forest_2007; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there is indication that CUE or $ANPP_{woody-stem}$/$GPP$ increase with latitude (Fig. S3). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.

>>>>

```{r setup, include = FALSE}
# knitr::opts_knit$set(root.dir = 'C:/Users/becky/Dropbox (Smithsonian)/GitHub/Global_Productivity/manuscript/')
knitr::opts_knit$set(root.dir = '/Users/kteixeira/Dropbox (Smithsonian)/GitHub/ForC-db/Global_Productivity/manuscript/')
```

@@

Carbon cycling in forests worldwide provides the energetic basis for sustaining the majority of Earth's terrestrial biodiversity and many human populations (**REF**), while strongly influencing atmospheric CO~2~ and climate [@bonan_forests_2008]. Forests' autotrophic carbon fluxes (FACF)--that is, carbon fixation, allocation, and metabolism by trees and other primary producers--sets the energy ultimately available to heterotrophic organisms (including microbes), in turn influencing their abundance (**REFS**) and possibly diversity [@waide_relationship_1999] (**REFS**). FACF influences all organic matter stocks in forest ecosystems and is linked to cycling of energy, water, and nutrients (**REFS**) [@piao_forest_2010]. Critically, FACF also define forest interactions with Earth's changing climate. Over 69 Gt of CO~2~ cycle through Earth's forests each year [@badgley_terrestrial_2019], and in recent decades their net C sequestration (~2.4 Gt C yr^-1^) offset roughly 30% of anthropogenic fossil fuel emissions [@pan_large_2011]. As atmospheric carbon dioxide levels continue to rise, driving climate change, forests will play a critical role in shaping the future of Earth's climate [@cavaleri_urgent_2015; @rogelj_mitigation_2018]. However, our ability to draw general macroscopic conclusions regarding global variation in FACF with respect to climate has been limited in that these analyses often mix forests that vary in stand age, disturbance history, and/or management status; do not always sufficiently parse related variables; and typically consider only one or a few variables at a time.

@@

The recent development of the Global Forest Carbon database (ForC), which synthesizes multiple variables and including records of stand history [@anderson-teixeira_carbon_2016; @anderson-teixeira_forc_2018], opens up the possibility for a standardized analysis of global scale variation in multiple FACF and the principle climatic drivers of these patterns. In order to approach this broad topic, we simplify the major gaps in our knowledge to five broad questions and corresponding hypotheses (Table 1). First, we ask how FACF vary with latitude. We then test how these fluxes relate to MAT and MAP, and additionally how they respond to other, less well-studied, climate variables. Finally, we consider the relationship between FACF and seasonality, considering the role of seasonality in explaining variation in carbon fluxes, and the influence of climate on FACF standardized by growing season length. We address the above questions for nine carbon fluxes contained in ForC, allowing for an in-depth exploration of the effect of climate on FACF globally.

@@

Variation in allocation to component carbon fluxes along latitudinal gradients was explored for the following pairings: $GPP:NPP$, $ANPP:BNPP$, $ANPP_{foliage}:ANPP_{woody.stem}$, $ANPP_{foliage}:NPP$, $ANPP_{woody.stem}:NPP$, $ANPP:NPP$, and $BNPP:NPP$. For each set of paired fluxes, measurements taken at the same site and plot, and in the same year, were paired together, and the ratio of each pair of measurements calculated. The ratios were regressed against latitude and climate variables, using the linear model specified above. Cook's distance analyses were carried out for each of the models, and indicated that data from a few high-elevation sites were having a disproportionate influence on the regressions. To account for this, models were re-run using only data from sites $\le$ 1000m.

@@

In total, we analyzed 1228 records from nine FACF variables taken from forests that had experienced no major anthropogenic disturbances within the past 100 years. These records represented a total of 154 distinct geographic areas across all forested biogeographic and climate zones (Fig. 1, Table 2).

@@

All major carbon fluxes decreased linearly with latitude (Fig. 2; Table S2). Latitude was a strong predictor for many of the carbon fluxes--particularly the larger fluxes (Table S2). Specifically, latitude explained 64% of variation in GPP (n = 254, p<0.0001), 50% in NPP (n = 114, p<0.0001) and 45% in ANPP (n = 259, p<0.0001). The FACF that were most poorly predicted by latitude were $BNPP_{root-fine}$ ($R^2$=0.17) and $ANPP_{woody.stem}$ ($R^2$=0.18). For all FACF, the relationship with latitude was best predicted by the linear model.

@@

All FACF increased linearly with MAT, and we found no support for a saturation point of FACF with MAT (all p<0.05; Figs. 3-4, S4-S5, Table S2). As with latitude, MAT tended to explain more variation in the larger FACF ($GPP$, $NPP$, $ANPP$, $R_{auto}$) and $ANPP_{foliage}$ (all $R^2$> 0.4) than in subsidiary and belowground fluxes ($ANPP_{woody-stem}$, $R_{root}$, $BNPP_{root-fine}$; all $R^2$< 0.25).

@@

All fluxes, with the exception of $R_{root}$, showed a positive linear relationship with solar radiation (Figs. S4-S5, Table S2). Solar radiation explained a low proportion of variability in all FACF, explaining less than 20% of the variation in each flux, with the exception of $R_{auto}$ ($R^2$ = 0.26, p<0.05).

@@

Temperature seasonality was a significant predictor of FACF. $GPP$, $NPP$, $ANPP$, and $R_{root}$ exhibited a polynomial relationship with seasonality (all p<0.05; Figs. 4e, S6-7; Table S2). $ANPP_{foliage}$, $ANPP_{woody-stem}$ and $R_{auto}$ decreased linearly with temperature seasonality (all p<0.05; Figs. 4e, S6-S7; Table S2). Temperature seasonality was strongly correlated with annual temperature range, which was likewise a similarly strong predictor of FACF (Table S2). FACF were highest where temperature seasonality = 0, and at an annual temperature range of 15$^\circ$C or lower.

@@

When FACF were standardized by growing season length, correlations with growing season climate--including temperature, precipitation, solar radiation, and PET--were generally weak (Figs. S8-S9). Speficifally, $ANPP$ increased with growing season temperature ($R^2$ = 0.10, p<0.001) and precipitation ($R^2$ = 0.04, p<0.05). Similarly, $ANPP_{foliage}$ increased slightly with growing season temperature ($R^2$ = 0.16, p<0.01) and precipitation ($R^2$ = 0.09, p<0.05). Growing season solar radiation had a positive influence on $GPP$ ($R^2$ = 0.21, p<0.001), $NPP$ ($R^2$ = 0.21, p<0.001), $BNPP$ ($R^2$ = 0.16, p<0.001) and $BNPP_{fine.root}$ ($R^2$ = 0.12, p<0.01). Growing season PET had a positive influence on $GPP$ ($R^2$ = 0.15, p<0.01), $NPP$ ($R^2$ = 0.18, p<0.01), $BNPP$ ($R^2$ = 0.23, p<0.0001), $BNPP_{fine.root}$ ($R^2$ = 0.11, p<0.05), and $ANPP_{woody-stem}$ ($R^2$ = 0.06, p<0.05). {**Becky, please verify/ edit the following:** There were no other signifcant correlations between growing season length-standardized FACF (9 variables in Table 2) and growing season climate (**which variables?**)}.

@@

Our analysis of a large global database (ForC) reveals how autotrophic carbon fluxes in mature forests vary with latitude and climate on a global scale. We show that, across all nine FACF analyzed, C cycling decreases continually with latitude (*H1.1*; Fig. 2)--a finding that confirms multiple previous studies but contradicts the idea that productivity of temperate forests rivals that of tropical forests [@huston_global_2009]. FACF generally increase in proportion to one another (*H1.2*), with few differences in allocation detectable at this global scale (Fig. S2) and with component fluxes summing appropriately to larger fluxes (Fig. 2), indicating no major, systematic omissions or overestimations of flux components. However, climate explained lower proportions of variability among subsidiary C fluxes (*e.g.*, $ANPP_{woody}$, $BNPP_{fine.root}$, $R_{auto-root}$; Fig. 2; Table S2). Latitudianal variation in FACF is primarily attributable to temperature-related variables (*H3, H4*), particularly MAT (Figs. 3-4). Water availability is also influential, but generally of secondary importance across the climate space occupied by forests (Figs. 3-4). Temperature seasonality and growing season length are closely correlated with MAT and are strong predictors of FACF (*H4*; Figs. 4e-f, S2, S6-S7), though growing season length doesn't improve upon MAT as a predictor. Within the growing season, the influence of climate on C cycling is smaller but still significant for a number of FACF (*H5*; Fig. S9; Table S4). These findings resolve a number of open debates and clarify the big picture of how FACF vary with latitude and climate on a global scale.

Past studies have differed in their conclusions regarding the relationship between FACF and latitude or its correlates (Table 1, *H1*)--quite possibly because of lack of standardization with respect to stand age and disturbance history. Our findings indicate that, among mature, undisturbed stands, FACF are unambiguously highest in the tropical regions, and the relationship is approximately linear (Fig. 2). This contrasts with the suggestion that productivity of temperate forests is similar to that of tropical forests [@huston_global_2009]. Compared to tropical forests, the temperate forest biome has experienced more widespread anthropogenic disturbance and has a larger fraction of secondary stands [@potapov_mapping_2008; @poulter_global_2018], so analyses comparing across latitudinal gradients without controlling for stand age risk confounding age with biome effects. In addition, because carbon allocation varies with stand age [@delucia_forest_2007] (**Doughty et al. 2018: DOI: 10.1111/btp.12504; Anderson-Teixeira et al. 2013: doi: 10.1111/gcb.12194**), age differences may introduce systematic biases into analyses of FACF across latitude or global climatic gradients. For example, woody productivity tends to be higher in rapidly aggrading secondary stands than in old-growth forests, where proportionally more C is allocated to respiration (**Doughty et al. 2018: DOI: 10.1111/btp.12504**);[@delucia_forest_2007; @kunert_understanding_2019]. By constraining our analysis to stands ≥ 100 years old with no record of major recent disturbance, we have clarified the shape of the relationship between FACF and latitude.

**This paragraph needs work--including re-assessment of results, interpretation, writing**
We show that, with the exception of some differences in MAT-MAP interactions (Fig. 3), FACF are broadly consistent in their responses to climate drivers on the global scale, with no pronounced trends in C allocation among the variable pairs tested (Figs. 2, S3). Although variation in allocation has been observed along gradients of elevation [@moser_elevation_2011] and water availability [@newman_above-_2006]--along with non-climatic axes of stand age [@litton_carbon_2007], nutrient availability [@litton_carbon_2007; @gill_belowground_2016], and forest structure [@taylor_greater_2019]--little variation in relation to climate is apparent at the global scale. Our conclusion, then, is that hypothesized gradients in allocation along global climate gradients cannot currently be supported for mature forests, although data quantity and standardization is currently insufficent to rule out the possibility that such trends exist. There have been suggestions that tropical forests tend to have low carbon use efficiency ($CUE$= $NPP$/$GPP$=($GPP$-$R_{auto}$/$GPP$), which are based on observations of low $CUE$ in old-growth tropical forests relative to (mostly younger) extratropical forests [@delucia_forest_2007; @malhi_productivity_2012; @anderson-teixeira_carbon_2016], but our analysis suggests that these low values might more appropriately be attributed to the fact that these forests are old than to their tropical climate. Indeed, $CUE$ is known to decline with forest age [@delucia_forest_2007; @collalti_is_2019], but appears to be roughly independent of $GPP$ [@litton_carbon_2007]. Among our sites with relevant data, there are no signficant trends in allocation of $GPP$ to $NPP$ or of $NPP$ to $ANPP_{woody-stem}$ across gradients of latitude or climate (Fig. S3). Additional measurements with careful methodological standardization across a consistent set of mature forest sites will be necessary to determine whether any climate-driven gradients in allocation exist at the global scale.

**there seems to be a difference between GPP and other fluxes at higher precips (Figs. 3, 4b), but the only allocaiton ratio we consider is GPP:NPP. It would be interesting too consider some others. Perhaps precip-->soil fertility-->CUE (see Doughty et al. 2018). I need to come back to this.**
